---
title: "Estimating, Testing, and Simulating Abundance in a Mark-Recapture with `recapr`"
author: "Matt Tyers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(149)
```



[Abundance Estimates and Standard Errors](#Abundance Estimates and Standard Errors)

[Confidence Intervals](#Confidence Intervals) 

[Simulation via Random Draws and Plotting Discrete Distributions](#Simulation via Random Draws and Plotting Discrete Distributions)

[Hypothesis Testing and Power Calculation](#Hypothesis Testing and Power Calculation)

[Sample Size Recommendation](#Sample Size Recommendation)

[Tests of Consistency for the Estimator](#Tests of Consistency for the Estimator) 

[Partially and Completely Stratified Estimators](#Partially and Completely Stratified Estimators) 



<a name="Hypothesis Testing and Power Calculation"></a>

## Abundance Estimates and Standard Errors

In a simple two-event mark-recapture experiment, point estimates of abundance can be calculated according to the Chapman, Petersen, or Bailey estimators using `NChapman()`, `NPetersen()`, or `NBailey()`, given the sample sizes in the first and second events, and the number of individuals recaptured in the second event.  The Petersen estimator is the MLE, but the Chapman estimator generally performs better and is often recommended.  Both assume sampling without replacement in the second sampling event.  By contrast, the Bailey estimator assumes sampling with replacement in the second event.

The argument `n1` denotes the sample size in the first event, with `n2` denoting the sample size in the second event and `m2` denoting the number of marked (recaptured) individuals in the second event.

```{r}
library(recapr)
NChapman(n1=100, n2=150, m2=20)    # Abundance estimate
vChapman(n1=100, n2=150, m2=20)    # estimated variance
seChapman(n1=100, n2=150, m2=20)   # standard error
```

<a name="Confidence Intervals"></a>

## Confidence Intervals 

Confidence intervals can be generated for the Chapman, Petersen, or Bailey estimators using `ciChapman()`, `ciPetersen()`, or `ciBailey()`.  A point estimate is returned as well as two confidence intervals, one using a Walt-type Normal-distribution assumption, and the other calculated by means of bootstrapping.  The bootstrapping interval is likely to have much more appropriate coverage than the Normal interval.

```{r}
ciChapman(n1=100, n2=150, m2=20)
```

<a name="Simulation via Random Draws and Plotting Discrete Distributions"></a>

## Simulation via Random Draws and Plotting Discrete Distributions

Vectors of random draws can be generated for the Chapman, Petersen, or Bailey estimators using `rChapman()`, `rPetersen()`, or `rBailey()` from an assumed value of total abundance (`N`).  This may be useful for simulation.

A plotting function is also provided for vectors of discrete (non-continuous) data.  Abundance estimates are calculated from count data, with the result having a non-integer but also non-continuous support.  It is possible that `plotdiscdensity()` may be more appropriate than a kernel density plot and more illustrative than a histogram.

```{r,fig.width=7, fig.height=4}
draws <- rChapman(length=10000, N=1500, n1=100, n2=120)
plotdiscdensity(draws)  
```

<a name="Hypothesis Testing and Power Calculation"></a>

## Hypothesis Testing and Power Calculation 

Approximate p-values can be returned using `pChapman()`, `pPetersen()`, or `pBailey()`, which use many random draws to simulate a null distribution.  The null hypothesis abundance is specified in the `nullN` argument, along with sample sizes `n1` and `n2`.  The observed abundance estimate can be specified using `estN`, or else the number of recaptures can be used directly, as `m2`.  The alternative hypothesis can be specified using the `alternative` argument, as `alternative="less"`, `"greater"`, or `"2-sided"`.

In the example given below, the null-hypothesis abundance is 500, with 100 individuals observed in the first and second events, with 28 recaptures in the second event.

```{r,fig.width=7, fig.height=4}
output <- pChapman(nullN=500, n1=100, n2=100, m2=28, alternative="less")
output

plotdiscdensity(rChapman(length=100000, N=500, n1=100, n2=100))   # null distribution
abline(v=500, lwd=2, lty=2)              # Null hypothesis abundance plots as a dashed line
abline(v=output$estN, lwd=2, col=2)      # Observed (estimated) abundance plots as a red line
```

Power calculation are also done using random draws from an assumed true (alternative) distribution, given the sample sizes of both events.  In `powChapman()`, `powPetersen()`, or `powBailey()`, argument `nullN` specifies the abundance used by the null hypothesis and `trueN` specifies the assumed true abundance used for the power calculation.  The `n1`, and `n2` arguments give the sample sizes and `alternative` gives the direction of the alternative hypothesis, with `alpha` specifying the level of the test to use. 

In the example given below, the power is calculated for a one-tailed test of a null abundance of 500, assuming a true abundance of 400.  The test powers are then calculated and plotted for assumed abundances from 250 to 450.  If the true abundance is 325, a one-tailed test of $H_0: N \geq 500$ will have a power of roughly 90%.

```{r,fig.width=7, fig.height=4}
powChapman(nullN=500, trueN=400, n1=100, n2=100, nsim=1000)

Ntotry <- seq(from=250, to=450, by=25)
pows <- sapply(Ntotry, function(x)
  powChapman(nullN=500, trueN=x, n1=100, n2=100, nsim=1000))
plot(Ntotry, pows)  
```

<a name="Sample Size Recommendation"></a>

## Sample Size Recommendation

Given the value of the sample size of one event and a best-guess at the true abundance, a recommendation for the sample size in the other sampling event can be calculated from the desired confidence and relative accuracy with the methods of Robson & Regier using `n2RR()`.  Recommendations for sample size are provided for two scenarios: if the size of the other sample is as specified, and if the two sample sizes are equal, which is the most efficient for sampling. The example below gives the full output for all allowed values of confidence and relative accuracy.  Output can be simplified by providing values in the `conf` and `acc` arguments.

```{r}
n2RR(N=1000, n1=100)
```

Output can be simplified by providing values in the `conf` and `acc` arguments.

```{r}
n2RR(N=1000, n1=100, conf=c(0.9,0.95), acc=c(0.15,0.1,0.05))
```

Simulation is used in `plotn2sim()`, in which the relative accuracy is calculated for a range of sample size values for the second event, for given levels of confidence.  The values to plot for n2 can be set by `n2range` and `n2step`, giving the range endpoints and step size, respectively.

```{r,fig.width=7, fig.height=6}
plotn2sim(N=1000, n1=100)
```

Simulation is also used in `plotn1n2simmatrix()`, in which the relative accuracy is calculated for a range of sample size values for both sampling events for a given level of confidence, in this case 95%.

```{r,fig.width=7, fig.height=6}
plotn1n2simmatrix(N=1000, nsim=2000)
```

<a name="Tests of Consistency for the Estimator"></a>

## Tests of Consistency for the Estimator

For a simple Petersen-type estimator of abundance to be consistent, one of three conditions must be true: either there must be complete mixing in the time between events, or the probability of capture must be equal for all individuals in the first event, or the probability of capture must be equal for all individuals in the second event.  This is typically investigated by means of a set of $\chi^2$ tests for each condition, with failure to reject the null hypothesis of any of the tests indicating that a simple Petersen-type estimator is reasonable to use.  Otherwise, if the null hypotheses are rejected in all tests, a completely stratified or partially stratified (Darroch-type) estimator should be used.

The `consistencytest()` function is provided, in the instance that sampling in both events is spatially or temporally stratified, in such a way that movement between strata may occur.  Event 1 and 2 strata do not need to be the same.  Arguments `n1counts` and `n2counts` accept vectors of sample sizes in each event by strata.  Recaptures of marked individuals can be specified in one of two ways: either as vectors of strata membership in `m2strata1` and `m2strata2`, or as a matrix in `stratamat`.  Arguments `m2strata1` and `m2strata2` accept vectors of the first- and second-event stratum membership for each recaptured individual, with only entries of `1, 2, 3, ...` accepted.  Alternatively, `stratamat` accepts a matrix in which each cell represents the number of recaptures for each combination of event 1 and event 2 strata, with rows corresponding to event 1 strata and columns corresponding to event 2 strata.

In the example below, there were two strata in event 1 (with sample sizes of 284 and 199) and three strata in event 2, and 30 individuals were marked in stratum 1 of event 1 and were recaptured in stratum 1 of event 2. These tests yield strong evidence of incomplete mixing in the time between events and of unequal capture probabilities in the first event, but also no evidence of unequal capture probabilities in the second event.  Therefore, use of a simple Petersen-type estimator may be considered justified.

```{r}
mat <- matrix(c(30,15,1,0,22,15), nrow=2, ncol=3, byrow=TRUE)
consistencytest(n1counts=c(284,199), n2counts=c(347,3616,1489), stratamat=mat)
```

In the event that there is no movement observed between strata, or the population is stratified in such a way that there can be no movement between strata (such as by age or sex), a stratified estimator should be used.  The `strattest()` function provides a means of conducting the typical $\chi^2$ tests to investigate the necessity of a stratified estimator.

Since strata membership remains the same in both events, usage is simpler, with `n1`, `n2`, and `m2` accepting vectors of first- and second-event sample sizes and recapture numbers by strata.  In the example below, there is strong evidence of unequal capture probabilities in the first event but not in the second.

```{r}
strattest(n1=c(100,100), n2=c(50,200), m2=c(20,15))
```

<a name="Partially and Completely Stratified Estimators"></a>

## Partially and Completely Stratified Estimators

If sampling is stratified in both sampling events and some movement exists between strata but mixing is incomplete in the time between events and capture probabilities are unequal in both events, a partially stratified (Darroch-type) estimator must be used.  The `NDarroch()` function provides a means for doing this, and argument usage is the same as that in `consistencytest()`.

The function returns abundance estimates and standard errors for each stratum in both sampling events, as well as an overall abundance estimate and standard error.

```{r}
mat <- matrix(c(59,30,1,45,280,38,0,42,25), nrow=3, ncol=3, byrow=TRUE)
NDarroch(n1counts=c(484,1468,399), n2counts=c(847,6616,2489), stratamat=mat)
```

If no movement can occur between strata, a completely stratified estimator may be used.  Functions `Nstrat()`, `vstrat()`, `sestrat()`, and `cistrat()`, are provided to compute abundance estimates, estimated variance, standard error, and confidence intervals,, and random draws for simulation.  In all cases, `n1`, `n2`, and `m2` accept vectors of first- and second-event sample sizes and recapture numbers by strata, and the additional `estimator` accepts the type of Petersen-type estimator to use: either `"Chapman"`, `"Petersen"`, or `"Bailey"`.  Additionally, `rstrat()` generates random draws from given vectors of `N`, `n1`, and `n2`.

```{r,fig.width=7, fig.height=5}
Nstrat(n1=c(100,200), n2=c(100,500), m2=c(10,10))
vstrat(n1=c(100,200), n2=c(100,500), m2=c(10,10))
sestrat(n1=c(100,200), n2=c(100,500), m2=c(10,10))
cistrat(n1=c(100,200), n2=c(100,500), m2=c(10,10))
draws <- rstrat(length=10000, N=c(500,1000), n1=c(100,200), n2=c(100,500))
plotdiscdensity(draws)
draws <- rstrat(length=100000, N=c(5000,10000), n1=c(500,200), n2=c(500,200))
plotdiscdensity(draws)
```

